Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques

### 2019-06-03 paper/paperV1.tex
JFP proofread corrections

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004, Ecole Centrale de Nantes, CNRS, F-44300, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
\maketitle

@@

Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} and traffic noise, human voices and birds calls can respectively be used as a proxy of its three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.\\

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
\end{enumerate}

@@

%In urban areas, the decrease of quality of sound environments resulting from urbanization is a major cause of annoyance, and has been identified as an important health factor. Consequently, the 2002/49/CE European directive~\cite{ec2002} requires the development of noise maps in large cities. These maps aim at providing useful tools for urban planning and noise reduction purposes. As an alternative to the noise-only paradigm, the soundscape approach defined in the ISO 12913-1~\cite{iso2014} norm as "the acoustic environment as perceived and understood and/or experienced by people and/or society, in context" extends the characterization of sound environments by considering the positive influence of some of its components.\\
%
%The assessment of soundscape quality through standardized perceptual descriptors has been extensively studied~\cite{viollon2000, axelsson2010, cain2013, jeon2018, aletta2016}. Specifically, the axis of pleasantness emerges as the first dimension in subjective assessments of soundscapes, and is thus increasingly associated with soundscape quality in the recent literature~\cite{axelsson2010, aumond2017, delaitre2014}. Pleasantness is consistently modeled as a function of both the perceived sound level and the activity of sources of interest within urban environments. While the exact taxonomy of sources used differs across studies, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010}, which can be simplified as traffic, human voices and birds~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance, emergence or time of presence.\\
%
%Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps. Leveraging the advent of the Internet of Things (IoT) and the availability of low-cost sensor networks~\cite{ardouin2018, mydlarz2017} could allow us to predict pleasantness or other perceptual factors from acoustic measurements at a reasonable cost. Though, bridging the gap between raw data issued from the sensors to reasonable estimates of perceptual attributes is yet an ongoing area of research.\\
%
%Ricciardi et al.~\cite{ricciardi2014} present a physical model of pleasantness using only variations of the measured sound level. Though, because these indicators are not representative of the composition of the sound scene, the model underperforms in comparison to predictions using perceptual assessments. Aumond et al.~\cite{aumond2017} explore new indicators based on spectral variations in recordings of sound scenes. These indicators are designed to model the activity of specific sound sources, specifically human voices and birds, but their discriminative properties are limited. In order to tackle this issue, we believe that the recent advances in machine learning can be useful to better approximate the contributions of sound sources relevant for perceptual attribute prediction such as pleasantness.\\
%
%In this paper, we do so by developing new indicators relying on source recognition models which are built based on deep learning techniques, which is a growing interest in the machine listening community~\cite{cakir2015, salamon2017-2}. Large amounts of annotated data are needed for these architectures to learn to extract relevant information from recordings and to the best of our knowledge such databases do not currently exist in the literature. We propose in this paper to study the potential of the use of simulated corpora~\cite{rossignol2015, salamon2017} for such purpose.\\
%
%More precisely, the contributions of this paper are three-fold:
%\begin{enumerate}
%  \item Provide controlled acoustic data and corresponding perceptual attributes relevant to the study of urban soundscapes perception, made available for the community.
%  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the time of presence of sound sources from raw acoustic data
%  \item Provide a comparative study of the behavior of the proposed approach with state of the art ones.
%\end{enumerate}
%
%First, a perceptual test is conducted in Section~\ref{sec:data} where simulated sound scenes are compared to recordings to validate their use for the prediction of pleasantness. Next, Section~\ref{sec:methods} presents the design of a deep learning architecture that performs multi-label sound source recognition at relevant time scales. Estimations of pleasantness using these predictions are next compared to existing models from physical indicators in Section~\ref{sec:results}.

\section{Data}
\label{sec:data}

@@

A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings, 45 seconds of audio in a single channel are extracted that represent well the properties of their respective ambiances, without single events overwhelming their overall perception. The resulting 45 second extracts are replicated to study perceptual differences between recorded and simulated scenes. This scale of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments. The recordings are first annotated in terms of the following properties:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and are associated with a sound level.
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

Sound scenes are then simulated using simScene~\cite{rossignol2015}, a software library in Matlab which allows the simulation of sound scenes as the additive composition of sound sources using a database of recordings associated to these sources. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB to 79.4~dB.\\

@@

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}. We choose to restrict the scene composition to these 3 sources. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance:

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios in dB for all events and backgrounds besides the first, in terms of normal distributions,
\item The inter-onset of event occurrences in seconds, also in terms of normal distributions.
\end{itemize}

Adjustments are made on the variance of the considered properties to better cover possible scenarios. The EBR of voice events, mostly read English recordings, are reduced to improve realism. An additional \textit{place} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions. For each scene the time of presence for the traffic, voice and bird sources is estimated using indicators proposed in~\cite{gontier2018}, and 75 scenes are chosen to maximize pairwise distance in the resulting 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned on the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB to 77.1~dB.\\

@@

A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point (0-10) scales. These scales are presented in french and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
\end{itemize}

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}

Prior to the test, participants are given a short verbal introduction and the interface is introduced before the test to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to attenuate ordering effects for the remaining of the test. Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\

@@

A total of 23 students aged from 22 to 23 including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing.

@@

Participants evaluate each scene on criteria using 11-point Likert scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the judgements of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus. Starting with the assessments of each participant as separate clusters, new clusters are formed iteratively where the average euclidean distance between all pairs within the two merged clusters at each step is minimal. and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\

\begin{figure}[!h]
    \centering
    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
\end{figure}

The perceptual responses of the 6 recorded and corresponding replicated scenes are then compared. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Kolmogorov-Smirnov statistical tests are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. Pleasantness is found to differ between the two groups only for one location, P4, which can be explained by high discrepancies in the perceptual time of presence of traffic and voices. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference in assessed time of presence of voices in the location P3, but as voices have a less distinct influence on pleasantness this is not reflected directly. Discrepancies can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes is found for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.\\

\begin{table}[h]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Kolmogorov-Smirnov test are in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & -1.04 & 0.43 & 0.13 & -0.91 & 0.39 & -2.09 & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & \textbf{0.91} & 0 & \textbf{-1.83} & 0.48 & 1.30 & -0.96 & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & -0.91 & 0.09 & \textbf{-1.43} \\
	P15 & -1.35 & 0.52 & 0.52 & -1.17 & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & 1.13 & -0.30 & -1.17 & -0.43 & 1.39 & -1.04 & -1.83 & 0.83 & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table}

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses at the scene level. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (left, n=25) and new scenarios (right, n=75). The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two corpora. For both sets the variance explained by the first two components is similar (resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%). Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, assessments for each scene are projected onto the PCA space and represented in Figure~\ref{fig:pspace_rec} as dots. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented by ellipses for one scene. The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/pca_p1.eps}
    \caption{Principal components analysis of the four general questions for scenes with real scenarios at the assessment level (n=575). Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown by ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.8\textwidth]{figures/pca_sim.eps}
    \caption{Principal components analysis of assessments for the simulated scenes (red dots) at the assessment level (n=1725), with projections of recorded and replicated scenes (blue dots).}\label{fig:pspace_sim}
\end{figure}

@@

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level. Typically this includes indicators derived from sound level measurements. For this study the following are computed with 1~s (slow) measurements using the Matlab ITA-toolbox~\cite{itatoolbox2017}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
\end{itemize}

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{fancyhdr}
\usepackage{a4wide} % Smaller margins = more text per page.
\usepackage[hidelinks]{hyperref}
\usepackage{graphicx}
\usepackage{epstopdf}
\usepackage{pdflscape}
\usepackage{multirow}
\usepackage{subcaption}
\usepackage{bbm}
\usepackage{color}
\DeclareMathOperator*{\argmax}{arg\,max}
\newcommand{\ml}[1]{\textcolor{red}{ML : #1}}
\pagestyle{fancy}
\begin{document}
\author{F\'elix Gontier $^1$, Catherine Lavandier $^2$, Pierre Aumond $^{2, 3}$\\Mathieu Lagrange $^1$, Jean-Francois Petiot $^1$}
\date{
$^1$ LS2N, UMR 6004 CNRS, Ecole Centrale de Nantes, F-44321, France\\
$^2$ ETIS, UMR 8051, University of Paris Seine, University of Cergy-Pontoise, ENSEA, CNRS, F-95000, France\\
$^3$ IFSTTAR, CEREMA, UMRAE, F-44344, Bouguenais, France
}
\title{Estimation of the perceived time of presence of sources in urban acoustic environment using deep learning techniques}
\maketitle

@@

Pleasantness is mainly linked to the perceived loudness~\cite{blauert1997, jekosch2004} but it also depends on the activity of sources of interest~\cite{nilsson2007, perez2012}. While the exact taxonomy of sources used differs across studies~\cite{guastavino2007, gygi2007, salamon2014, brown2011}, three major source types are found to contribute to pleasantness in urban contexts: technological, human and natural sources~\cite{nilsson2007, axelsson2010} where traffic noise, human voices and birds calls can respectively be used as a proxy of these three types~\cite{lavandier2006, ricciardi2014, aumond2017}. In existing models, source activity can be quantified by the dominance~\cite{hong2017}, the emergence~\cite{guastavino2006} or the time of presence~\cite{ricciardi2014}. Though, the perceptual characterization of soundscapes requires subjective inputs, and thus cannot directly be applied to the production of perceptually motivated noise maps.\\

@@

More precisely, the contributions of this paper are three-fold:
\begin{enumerate}
  \item Provide a corpus of simulated scenes for which relevant acoustic properties are well controlled.
  \item Propose numerical means~\footnote{Code is open source and available at the companion website.} for predicting the perceived time of presence of sound sources from raw acoustic data using deep learning approaches.
  \item Demonstrate through a comparative study with state of the art approaches that the proposed method allows the prediction of pleasantness to a sufficient degree.
\end{enumerate}

@@

\section{Perception of simulated sound scenes}
\label{sec:data}

@@

A corpus is first created to validate the use of simulated sound scenes in the development of time presence prediction models. It is based on recordings from one of the four soundwalks performed in~\cite{aumond2017}, including 19 locations in the 13th district of Paris, France. A classification of these scenes is proposed in~\cite{gloaguen2017} in terms of ambiance: \textit{park}, \textit{quiet street}, \textit{noisy street} and \textit{very noisy street}. For each of the 19 recordings, 45 seconds of audio in a single channel are extracted. This duration of 45 seconds is chosen to be perceptually relevant for the retrospective evaluation of sound environments. The 45~s segments are selected to represent the properties of their respective ambiances in terms of source composition, without single events overwhelming their overall perception. The extracts are then subjectively annotated in terms of the following properties:

\begin{itemize}
\item Background sources that are present throughout the whole scene, and their respective sound level considered constant.
\item Sound events characterized for each occurrence by their source type, onset-offset and event-to-background ratio (EBR) in dB.
\end{itemize}

The 45~s sound scenes are then replicated using \textit{simScene}~\cite{rossignol2015}. \textit{simScene} is a software library in Matlab that allows the simulation of sound scenes as the additive composition of sound sources, using a database of recordings associated to these sources. Here the database is constructed from excerpts of the LibriSpeech~\cite{panayotov2015} corpus for voices and Freesound\footnote{\url{https://freesound.org}} contributions for remaining sources. The original recordings from 6 locations (P1, P3, P4, P8, P15 and P18) are chosen for perceptual comparison, as they explore diverse real-life situations with respect to environment categorization. All 6 recorded and 19 replicated scenes are normalized so that their playback sound level through the restitution system is the same as measured during recording, with a range from 63.9~dB SPL to 79.4~dB SPL.\\

@@

To further study the perceptual properties of simulated scenes, the corpus is extended with 75 simulated scenes corresponding to new scenarios. Previous studies show that pleasantness is mainly influenced by the activity of three sources: \textit{traffic}, \textit{human voices} and \textit{birds}. We choose to restrict the scene composition to these 3 sources. Newly generated scenarios should cover most real-life situations while remaining perceptually plausible. Thus, statistics on background and event properties are extracted for each of the three sources from annotations presented in~\cite{gloaguen2017}. These statistics include, conditionally to the ambiance (resp. \textit{park}, \textit{quiet street}, \textit{noisy street}, \textit{very noisy street} and \textit{square}):

\begin{itemize}
\item The probability of appearance of a given source in the scene, for which events and backgrounds are considered separately,
\item Event-to-background ratios (EBR) in dB for all events and backgrounds compared to a main background source, drawn from gaussian distributions with parameters identified from annotations of recordings,
\item The inter-onset of event occurrences in seconds, also drawn from gaussian distributions.
\end{itemize}

Adjustments are made on the variance of the considered properties to better cover possible scenarios. Voice events consist mostly of read English recordings, thus their EBR is reduced to improve realism. An additional \textit{square} ambiance is added with properties derived empirically from available data. Diverse new scenarios are then generated by sampling the background and event properties distributions, resulting in 40 scenes per ambiance for a total of 200 scenes. For each scene the perceptual time of presence for the traffic, voice and bird sources is estimated using the indicators proposed in~\cite{gontier2018}. Ideally, sound scenes should cover the resulting 3-dimensional space in a homogeneous way. Though, the time of presence is not a parameter of the simulation process and thus the distribution of the scene population in this space is not controlled. As an alternative, 75 scenes are selected that maximize the minimum pairwise distance in the 3-dimensional space. This is shown in Figure~\ref{fig:tvb_pres}. The playback sound level is also conditioned to the ambiance according to typical sound levels in urban context, and ranges from 46.6~dB SPL to 77.1~dB SPL.\\

@@

A perceptual experiment is conducted where participants are asked to evaluate sound scenes in terms of 9 criteria represented by 11 point scales (0-10). These scales are presented in French and translated in this report using standard terminology. The first 5 scales relate to general properties of the scene:
\begin{itemize}
\item Pleasantness: \textit{Unpleasant - Pleasant} (\textit{D\'esagr\'eable - Agr\'eable}) - P
\item Liveliness: \textit{Inert, amorphous - Lively, eventful} (\textit{Inerte, amorphe - Anim\'e, mouvement\'e}) - L
\item Overall loudness: \textit{Quiet - Noisy} (\textit{Silencieux - Bruyant}) - OL
\item Interest: \textit{Boring, uninteresting - Stimulating, interesting} (\textit{Ennuyeux, inint\'eressant - Stimulant, int\'eressant}) - I
\item Calmness: \textit{Agitated, chaotic - Calm, peaceful} (\textit{Agit\'e, chaotique - Calme, tranquille}) - C
\end{itemize}

@@

The perception of source activity can be linked to the notions of dominance, time of presence, or emergence. We choose the perceptual time of presence, that is the percentage of time for which the corresponding source is heard, as the activity descriptor for this study. The influence of traffic on pleasantness is further separated into that of background and salient events. Thus, 4 additional questions are presented to the participants and evaluated on the same 11 point scales:
\begin{itemize}
\item Sound level of passing vehicles: \textit{Very low - Very high} (\textit{Tr\`es faible - Tr\`es fort}) - $L_{T, p}$,
\item Time of presence of Traffic, Voices, Birds: \textit{Never - Continuously} (\textit{Jamais - Continuellement}), resp. $T_{T, p}$, $T_{V, p}$ and $T_{B, p}$
\end{itemize}
where $p$ denotes a perceptual evaluation.\\

Prior to the test, a short verbal introduction is given to the participants and the interface is introduced to ensure that the quantities are well understood. Although the corpus is comprised of 100 sound scenes, participants only evaluate 50 scenes: all listen to the 6 recorded and 19 replicated sound scenes, then to 25 of the 75 simulated with new scenarios according to a balanced incomplete block design~\cite{dagnelie2003}. The selection of simulated scenes is done so that all scenes in the sub-corpus are evaluated by the same number of participants. All participants are first presented with the most quiet then most noisy of the recorded scenes (resp. P3 and P15). A random listening order is generated for each subject to control ordering effects for the remaining of the test. Participants can listen to each scene once, and have to answer all questions and listen to the full extract before proceeding to the next.\\

@@

A total of 23 students aged from 22 to 23 yo including 16 males and 7 females at Ecole Centrale de Nantes completed the test. All gave consent prior to the experiment and reported normal hearing. With 23 subjects, the incomplete block design is not perfectly balanced.

@@

%Participants evaluate each scene on criteria using 11-point Likert scales. Many statistical tools assume that the studied data follows continuous, single mode distributions. It is thus important to first verify that no distinct groups of participants can be obtained from perceptual assessments. A hierarchical clustering analysis of the judgements of the participants is performed for the 25 scenes in the commonly evaluated sub-corpus. Starting with the assessments of each participant as separate clusters, new clusters are formed iteratively where the average euclidean distance between all pairs within the two merged clusters at each step is minimal. and shown in Figure~\ref{fig:hclusters}. No clear separation into groups is found, as a result the assessment distributions are assumed to be single-mode and their means can be used for further analysis.\\
%
%\begin{figure}[!h]
%    \centering
%    \includegraphics[width=0.7\textwidth]{figures/subj_gr.eps}
%    \caption{Dendrogram of the hierarchical clustering of participants on the first 25 scenes of the test. No distinct groups are visible.}\label{fig:hclusters}
%\end{figure}

The perceptual responses of the 6 recorded and corresponding replicated scenes are first compared. Table~\ref{tab:ogrep} shows the mean differences between assessments for pairs of scenes with equivalent scenarios. Wilcoxon signed-rank tests~\cite{wilcoxon1945} are implemented for each scene and perceptual scale to outline significant differences between assessment distributions, which are shown in bold. As the data is discretely distributed, zero differences between paired samples are included using Pratt's modification of the test~\cite{pratt1959}. On the general scales (P, L, OL, I and C), all mean differences are lower than 2 points on the 11-point Likert scale (catherine?). Though, significant differences are outlined that can be linked to corresponding discrepancies in source-specific parameters. The highest difference (-5.22) is found for the assessment of the time of presence of traffic in the location P4. For this location, the background traffic in the recorded scene varies along time, it is louder in the first half of the scene than in the second half. Replicating this scene using simScene imposes a constant sound level for background sources. Thus, the background traffic is louder in the replicated scene than it is in the recording for about half of its duration. To a lesser extent the same issue explains the large difference (-4) in the assessed time of presence of voices in the location P3. Discrepancies on source-specific scales can also be interpreted by the choice of isolated samples, which is semi-random and based on a high-level source taxonomy. For example, no difference is made during annotation between child or adult speech, or depending on its expressiveness. Overall, no consistent difference between the perception of recorded and replicated scenes emerges for the studied points. Thus, the proposed hypothesis that recorded and replicated scenes generate the same perceptual responses cannot be rejected.\\

\begin{table}[h]
\centering
\caption{Mean differences of perceptual assessments between original and replicated sound scenes. Significantly different distributions as per a Wilcoxon signed-rank test are shown in bold (n=23, p<0.05)}
\label{tab:ogrep}
%\resizebox{\columnwidth}{!}{
\begin{tabular}{ c | c c c c c c c c c }
\hline
	 & P & L & OL & I & C & $L_{T, p}$ & $T_{T, p}$ & $T_{V, p}$ & $T_{B, p}$ \\ \hline
	P1 & 0.43 & \textbf{-1.65} & \textbf{-1.04} & 0.43 & 0.13 & \textbf{-0.91} & 0.39 & \textbf{-2.09} & 0.61 \\
	P3 & 0.26 & -0.43 & 0.30 & -1 & 0.30 & 0.35 & 1.04 & \textbf{-4} & 0.22 \\
	P4 & 0.91 & 0 & \textbf{-1.83} & 0.48 & \textbf{1.30} & \textbf{-0.96} & \textbf{-5.22} & \textbf{1.43} & 0.04 \\
	P8 & 0.26 & \textbf{-1.65} & -0.87 & -0.96 & 0.65 & \textbf{-2.04} & \textbf{-0.91} & 0.09 & \textbf{-1.43} \\
	P15 & \textbf{-1.35} & 0.52 & 0.52 & \textbf{-1.17} & 0.09 & 0.61 & 0.13 & \textbf{1.96} & \textbf{-2.74} \\
	P18 & \textbf{1.13} & -0.30 & \textbf{-1.17} & -0.43 & \textbf{1.39} & -1.04 & \textbf{-1.83} & \textbf{0.83} & \textbf{1.30} \\ \hline
\end{tabular}
%}
\end{table}

To validate the use of simulated sound scenes with new scenarios as well as reduced source complexity in the development of prediction models, the perceptual space generated by the experiment's five general scales (pleasantness, liveliness, overall loudness, interest and calmness) is studied. It is obtained by performing a principal components analysis (PCA) on the corresponding perceptual responses averaged along participants. No standardization is applied to the data. Figure~\ref{fig:pspace_rec} and Figure~\ref{fig:pspace_sim} compare the results for scenes based on recordings (n=25) and new scenarios (n=75) respectively. The resulting spaces are similar, with only overall loudness and pleasantness axes slightly rotated between the two subcorpora. For both sets the variance explained by the first two components is similar, resp. 79.4\% - 18.1\% and 79.6\% - 15.2\%. Furthermore, these representations are comparable to those found in previous work on perceptual dimensions~\cite{axelsson2010, cain2013}. Thus, the use of simulated scenes based on both real and new scenarios does not result in major differences in the relations between perceptual quantities. Additionally, the assessments averaged on all subjects are projected onto the PCA space for each scene and represented in Figure~\ref{fig:pspace_p1} as red dots for simulated scenes corpus and blue dots for recorded and replicated scenes. The space covered by scenes based on new scenarios covers that of the studied real-life environments, further demonstrating the effectiveness of the scene generation procedure in terms of diversity of scenarios.\\

Discrepancies between projections of original and replicated scenes are highlighted in Figure~\ref{fig:pspace_rec} using arrows. The standard deviation of assessments is represented by ellipses for one location (P1). The projections of assessment distributions consistently overlap for all pairs of recorded and replicated scenes, available on the companion website~\footnote{URL}. This illustrates the results in Table~\ref{tab:ogrep} where perceptual assessments were not found to differ significantly overall.\\

\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/pca_p1.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 6 recorded and 19 replicated scenes (n=25). Red dots are projections of the assessments of each scene averaged over subjects. Arrows show discrepancies between corresponding recorded and replicated scenes. For the P1 recording location the standard deviation of the projection of individual responses is shown by ellipses.}\label{fig:pspace_rec}
\end{figure}
\begin{figure}[h]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/pca_sim.eps}
    \caption{Biplot of the principal components analysis of average assessments for the 5 general questions on the 75 simulated scenes (n=75). Assessments of simulated scenes are projected as red dots, and recorded and replicated scenes as blue dots.}\label{fig:pspace_sim}
\end{figure}

@@

First, indicators can be computed directly from the mixed audio, without the need for information about the composition of the scene, \textit{i.e.} source activity and level. Typically this includes indicators derived from sound level measurements. For this study the following variables are computed with a time frame of 1~s using the Matlab ITA-toolbox~\cite{itatoolbox2017}:

\begin{itemize}
\item Z-weighted $L_{eq}$ and A-weighted $LA_{eq}$ equivalent sound levels in dB and dBA respectively.
\item $L_{10}$, $L_{50}$ and $L_{90}$: 10th, 50th and 90th percentiles of the Z-weighted sound level. The $L_{10}$ is often associated to events and the $L_{90}$ to background activity, while the $L_{50}$ is a measurement of the overall sound level.
\item $LA_{50}$: 50th percentile of the A-weighted sound level, with similar properties as the $L_{50}$.
\item $L_{50, 1kHz}$: 50th percentile of the Z-weighted sound level for the 1kHz frequency band, also a good descriptor of the overall sound level of the scene.
\item $LA_{10}-LA_{90}$: Emergence indicator included in the pleasantness model presented in~\cite{ricciardi2014}.
\end{itemize}

The time and frequency second derivative ($TFSD$) is introduced in~\cite{aumond2017} as a descriptor of source activity. Its expression is:
\begin{equation}
TFSD_{f, t} = \frac{\lvert\frac{d^2L(f, t)}{dfdt}\rvert}{\sum_{f1=31.5Hz}^{f1=16kHz}\lvert\frac{d^2L(f, t)}{dfdt}\rvert}
\end{equation}
where $L(f, t)$ is the third-octave spectrum of the signal. To provide a baseline for pleasantness prediction, the $TFSD$ is computed for the 4kHz band and 125~ms measurements ($TFSD_{4kHz(1/8s)}$) to represent birds, and for the 500Hz band with 1~s measurements ($TFSD_{500Hz(1s)}$) to represent voices.\\
